Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-09-08 Manuscript/manuscript.tex
Modifcations after Hectors' comments.

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
like in signal propagation in neurons and vesicle fusion.
\todo{JOE: list more introductory examples, where it has interesting effects.}
Zwitterionic phosphocholine (PC) lipid bilayers have been widely used as model systems
to understand molecular level details of specific cation interactions with
cellular membranes by using experimental \cite{cevc90,tocanne90,binder02,pabst07,uhrikova08}
and theoretical methods \cite{bockmann03,bockmann04,Berkowitz12,??} 
\todo{PAVEL: introduce previous theoretical work that discusses cation binding to POPC w.r.t. its specific moiteties, e.g. Lukas' paper}.
While relative binding affinity of different ions is agreed to follow Hoffmeister
series, the molecular details of binding and binding energetics are
not agreed in the literature \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Non-invansive spectroscopic methods, like nucelar magnetic resosnance (NMR), scattering
and infrared scpectroscopy, give accurate information about ion binding
in lipid bilayers\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02}. 
Molecular level interpretation of the results, however, requires assumptions
about binding models and is often not fully conclusive \cite{??}.
On the other hand, the accuracy of the state of the art atomistic resolution lipid and ion models
have turned out insufficient for the detailed interpretation of cation binding details \cite{catte16}.
The overestimated specific cation binding observed in current lipid models \cite{catte16}
may lead to articifially positively charged memranes and significant artefacts in
MD simualtions (e.g.~divalent cation-induced charge inversion of bacterial membranes~\cite{Luan2016}
\todo{JOE: this may be a too daring statement. \\
  SAMULI: I think that the divalent cation induced charge inversion is not an artefact,
  but it might happen with too small concentrations. 
  On the other hand, it is not a new observation either, see e.g. literature cited in NMRlipids IV project.}).

@@

In this work we show that the accuracy of ion binding, especially calcium,
in zwitterionic PC lipid bilayer can be significantly improved 
for classical MD simulation models of zwitterionic lipids
by including electronic polarizability
in the polar region of lipid molecules
using the electronic continuum correction (ECC)~\cite{leontyev11}. 
This approach has been previously shown to improve performace of
ion models against neutron scattering data in concentrated solutions 
and ab-initio simulations~\cite{Jungwirth2017,Jungwirth2015,kohagen14,kohagen16}. 
The realistic structure of ions in a bulk solvent was not, however, sufficient to correct binding in lipid
bilayers~\cite{catte16}. 
To fix it, we propose
a classical fixed-charge MD simulation model of 
1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) lipid 
that accounts for electronic polarizability by using ECC. 
We will validate such a model by comparing its lipid bilayer structue
to x-ray scattering data and its head group structure to NMR experiments. 
We will then quantify the accuracy of the head group order parameter response to the bound charge
and compare it to the experimental NMR data of sodium and calcium binding in PC lipid bilayers. 
In addition, we will quantify relative binding affinities of different moieties in POPC to such cations. 
This will be discussed in terms of binding modes, stoichiometry and binding isotherms and their models used in experiments.
The proposed POPC model, ECC-lipids, and the approach to obtain it
is, hence, highly useful for future MD simulations with physiological salt conditions, 
and for effects where membrane electrostatics play a major role.

@@

\subsection{Electrometer concept} \label{section:electrometer}
Ion binding was compared between experiments and simulations by 
using lipid head group order parameters and 
''electrometer concept'' \cite{seelig87,catte16}.
The concept is based on the experimental observation that the C-H bond
order parameters of $\alpha$ and $\beta$ carbons in lipid head group
(see Fig.~\ref{simVSexpNOions}) are proportional to the amount 
of bound charge in a lipid bilayer \cite{seelig87}.
This is rationalized as a change of lipid head group dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative charge \cite{seelig87}. 
More recent analysis in the NMRlipids project concluded 
that the order parameters decrease with bound positive charge 
and increase with bound negative charge when the signs are taken 
in account \cite{ollila16,catte16}. The concept was also shown
to be suitable for the comparison of ion binding affinity in lipid 
bilayers between simulations and experiments~\cite{catte16}.

The order parameters for $\alpha$ and $\beta$ segments in head group, as well as for
all C-H bonds in lipid molecules,
can be accurately measured by using $^2$H NMR or $^{13}$C NMR techniques.
The order parameters can be calculated from MD simulations 
by using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and average is taken over all sampled configurations \cite{ollila16}.

According to the electrometer concept,
the change of the head group order parameters as a function of
the amount of bound charge per lipid $X^\pm$ 
can be written as~\cite{ferreira16}
\begin{equation}\label{OPchangeEQ}
\Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) =\frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
%the empirically observed linear relation %for the change in order parameters %with respect to a bilayer without bound charges              
%\begin{equation}\label{electrometer_eq}                                                                                                   
%S_{\rm{CH}}^{i}(X^\pm)=S_{\rm{CH}}^{i}(0) + \frac{4m_i}{3\chi} X^\pm,                                                                     
%\end{equation}                                                                                                                            
$m_i$ is an empirical constant depending on the valency and position of bound charge,
and the value of the quadrupole coupling constant is $\chi \approx$\,167\,kHz.
Atomic absorption spectra and $^2$H NMR data gave
$m_\alpha=-20.5$  and $m_\beta=-10.0$ for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
Before using electrometer concept to analyze cation binding, 
we calibrate the response of head group order parameters to the bound
charge in the simulation models. This is done by comparing 
the slope $m_i$ in Eq. \ref{OPchangeEQ} between simulations and
experiments of cationic surfactants in POPC bilayer \cite{scherer89}. 
Since essentially all surfactants locate in bilayer, the 
molar ratio directly gives the amount of bound charge per 
lipid $X^\pm$ is these systems.

@@

In order to validate the newly developed model, ECC-lipids, 
we compared our simulation results without any ions to
NMR order parameters measurements and x-ray scattering form factors 
(Fig.~\ref{simVSexpNOions} and Table~\ref{tab:apls}). 
The tail order parameters being already highly optimized in the original Lipid14 model~\cite{dickson14}
are found to match the experimental values nicely.
The head group and glycerol backbone order parameters accuracy is 
comparable to the state of art lipid models available in literature \cite{ollila16}.
The agreement between the x-ray scattering form factors and the areas per molecule from simulations and experiments
confirm that the membrane structural properties are well captured. 
A structural comparison of ECC-lipids with Lipid14 can be found 
in SI along with results with other water models. 
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.

@@

The headgroup order parameter changes with increasing amount of 
cationic surfactant dihexadecyldimethylammonium bromide (C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) 
is compared between experiments~\cite{scherer89} and simulations in
Fig. \ref{OrderParameterCHANGESsurf}. 
Approximately linear decrease of the order parameters is observed
in simulations and experiments with increasing amount of cationic
surfactants up to mole fractions of $\sim$50\%,
as expected from Eq. \ref{OPchangeEQ}.
The slope is, however, too steep in Lipid14 model indicating that 
the head group order parameters are too sensitive to a bound charge.
The ECC-lipids model gives a slope in very good agreement with experiments
for the $\alpha$ segment, while the slope is slightly
underestimated for the $\beta$ segment.
%This has to be taken into account when interpreting the
%head group order parameter response to CaCl$_2$ concentration in Lipid14 model.
%observed in  Ref. \citenum{catte16}, see also Fig. \ref{OrderParameterCHANGESnewMODELS}.
%The present comparison of head group order parameter changes 
%suggests that the observed overestimated response of all models to 
%increasing CaCl$_2$ concentration in \cite{catte16} 
%can be explained at least in part by a high sensitivity of the head group order parameter response
%to the bound charge.
\todo{SAMULI: We could calculate the slopes from simulations, but I am not sure if
we would actually learn anything useful from this.}

The headgroup P-N vector angle with respect to the membrane normal
is also shown as a function of cationic surfactant mole fraction
in Fig.~\ref{OrderParameterCHANGESsurf}.
The headgroup orients more parallel to the membrane
normal with increasing amount of bound cations, as expected~\cite{seelig87}. 
The effect is more pronounced in Lipid14 than 
in ECC-lipids model, which is in line with the order parameter 
results and the reduced charge--dipole interactions in the ECC-lipid model.
The response of $\alpha$-order parameter to bound positive charge
in ECC-lipid model is in good agreement with experiments.
Therefore we use the data in Fig.~\ref{OrderParameterCHANGESsurf}
to suggest a linear relation between $\alpha$-carbon order parameter
and average P-N vector angle changes $\Delta \Theta_{\rm P-N}=?? \Delta S_{\rm CH}^{\alpha}$.
This can be used to estimate the change in P-N vector angle
from experiments measuring changes in $\alpha$-carbon order parameter. 
\todo{SAMULI: I think that we should establish the relation between 
order parameter and P-N vector angle changes by using ECC-lipids model.
This would be useful for people measuring aplha order parameter
changes from experiments.}
For example, $\Delta S_{\rm CH}^{\alpha}=0.018$ measured due to the
addition of 66.8 $\mu$M concentration of Mellittin in bulk water \cite{kuchinka89}
gives $\Delta \Theta_{\rm P-N} = 2.5^{\circ}$.
\todo{I looked this roughly from the figure, should be calculated from 
the actual relation if we decide to put this.}

@@

The binding affinities are quantified by using
the ion density profiles along membrane normal shown
in Fig.~\ref{fig:cacl-dens}. The density profiles show
larger Ca$^{2+}$ density peak in lipid headgroup region for
Lipid14 model with Dang and ECC-ions than for the ECC-lipid model.
The relative surface excess calculated from Eq.~\ref{surfexcess} gives 
$\Gamma_i^w = 0.07 \pm 0.01 nm^{-2}$ for the ECC-lipid model,
which is significantly smaller than $\Gamma_i^w = 0.13 \pm 0.01 nm^{-2}$ for
Lipid14 with \AA{}qvist and $\Gamma_i^w = 0.3 \pm 0.03 nm^{-2}$ with Dang ions.
Rough estimates for the free energy difference between bound and unbound cation
are given by
\begin{equation}
  \Delta G={\rm k_b T}\log(\frac{p_o}{p_i}),
\end{equation}
where $p_o$ and $p_i$ are estimated from the Ca$^{2+}$densities in bulk water and
in the maximum density in bilayer, respectively. The density profiles in
Fig. \ref{fig:cacl-dens} give $\sim 0.8$ k$_{\rm b}$T for the free energy difference
between bound and unbound Ca$^{2+}$ ions in ECC-lipid model and $\sim 1.4$ k$_{\rm b}$T
in Lipid14 with \AA{}qvist.

@@

Binding stoichiometry of \ce{Ca2+} and POPC was 
thoroughly studied in the experimental work~\citenum{altenbach84}.
Several binding models were proposed and tested of which only one,
ternary complex binding model, provided a good fit of the experimental observations. 
In such a model, it is assumed that \ce{Ca2+} cations bind to a POPC membrane 
with a stoichiometry 2~POPC:1~\ce{Ca^{2+}}.
Such a model also provides a good fit 
to our simulations with ECC-lipids 
(see Fig.~\ref{fig:cacl-bind} in SI and its caption for details). 
Simulations, however, allow us to directly evaluate the stoichiometry 
by calculating relative propensities of various \ce{Ca2+}:$n\times$POPC clusters
In Figure~\ref{fig:cacl_complexes} we see that ternary complex is indeed 
the most probable binding mode of calcium at 285~mM concentration.

Using our model, ECC-lipids, we fitted such a model to our simulations with a g
In such a model, there is a linear relationship between quantities 
$C_b$, mole fraction of bound \ce{Ca2+} per POPC, and $\sqrt(C_b/C_I)$, 
where $C_I$ is the concentration of free cations at the plane of ion binding~\cite{altenbach84}.
The concentration $C_b$ was obtained from an extrapolation of linear relation 
between deuterium NMR measurements and atomic absorption spectroscopy for low concetrations of \ce{CaCl2}.
Such an extrapolation is valid as long as the mode of \ce{Ca2+} binding 
remains constant throughout the extrapolation range. 
The concentration $C_I$ is determined by using the surface potential by using the Boltzmann equation.
However, Boltzmann theory yields inaccurate results
for divalent cations like \ce{Ca2+}~\cite{Andelmann1995}. 
An atomistic simulation, on the other hand, provides these quantities directly without severe assumptions. 
Hence we hypothesise that the discrepancy between the results in the experiment~\cite{altenbach84} and 
our simulations likely lays in the fact that 
the assumptions and relations used for determinig concentrations $C_b$ and $C_I$ in the experiment~\cite{altenbach84}
gradually do not hold for higher concentrations of \ce{Ca2+}.

Binding probability:
\ce{Ca2+} complex with
1 POPC - 13\%
2 POPC - 18\%
3 POPC - 12\%

\begin{figure}[tbp]
  \centering
  \includegraphics[width=6.0cm]{../Fig/number_bound_lipids_p_cation_Ecc-lipids.eps} \\
  \caption{\label{fig:cacl_complexes}
      Histogram of relative probabilities of existence of complexes
      of \ce{Ca2+} and $n$ POPC lipids. 
  }
\end{figure}

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=6.0cm]{../Fig/volmap_resid10_Ca_Cl_PO4Cent.png} \\
  \includegraphics[width=4.0cm]{../Fig/volmap_resid10_Ca_Cl_sn1Cent.png}
  \includegraphics[width=3.6cm]{../Fig/volmap_resid10_Ca_Cl_sn2Cent.png}
  \caption{\label{fig:volmaps}
      Contours of probability isodensities of \ce{Ca2+} with respect to 
      various moieties fixed in space (highlighted with transparent spheres): 
      phosphate moiety, side chain 1 carbonyl group and side chain 2 carbonyl group.
      Shown contours suggest that the dominant contribution 
      to \ce{Ca2+} binding likely comes from the phosphate group, 
      but remain inconclusive on the details of the interactions with any of the two carbonyl groups. 
  }
  \todo{JOE: I'll update this figure with some ensemble of configuration to support binding preference of \ce{Ca2+}}
\end{figure}

@@

It was found in the original work~\cite{altenbach84} that 
a ternary complex binding model (i.e.~2~POPC:1~\ce{Ca^{2+}})
provides the best fit to experimental measurements of all considered models in that study. 
In such a model, there is a linear relationship between quantities 
$C_b$, mole fraction of bound \ce{Ca2+} per POPC, and $\sqrt{C_b/C_I}$, 
where $C_I$ is the concentration of free cations at the plane of ion binding~\cite{altenbach84}.
The concentration $C_b$ was obtained from an extrapolation of linear relation 
between deuterium NMR measurements and atomic absorption spectroscopy for low concetrations of \ce{CaCl2}.
Such an extrapolation is valid as long as the mode of \ce{Ca2+} binding 
remains constant throughout the extrapolation range. 
The concentration $C_I$ is determined by using the surface potential by using the Boltzmann equation.
However, Boltzmann theory yields inaccurate results
for divalent cations like \ce{Ca2+}~\cite{Andelmann1995}. 
An atomistic simulation, on the other hand, provides these quantities directly without severe assumptions. 
Hence we hypothesise that the discrepancy between the results in the experiment~\cite{altenbach84} and 
our simulations likely lays in the fact that 
the assumptions and relations used for determinig concentrations $C_b$ and $C_I$ in the experiment~\cite{altenbach84}
gradually do not hold for higher concentrations of \ce{Ca2+}.

\begin{figure}[]
  \centering
  \includegraphics[height=9.0cm,angle=-90]{../Fig/bound-CAs_conc-eccl17.eps}
  \caption{\label{fig:cacl-bind}
    Ternary complex binding model of \ce{Ca2+} to a POPC membrane 
    that assumes the stoichiometry of 2~POPC:1~\ce{Ca^{2+}} (details in reference~\citenum{altenbach84}) 
    provides a good fit to experimental measurements~\cite{altenbach84}
    and it also provides a good fit to our simulation data. 
    Note that the units in the reference~\citenum{altenbach84} are different from the units presented here,
    and, hence, the observed slope of the linear relationship is different.
    }
\end{figure}

>>>>

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
like in signal propagation in neurons and vesicle fusion.
\todo{This could be maybe extended.}
Since the direct measurements of ion-membrane interactions from biological systems
are difficult, lipid bilayers are often used as models systems for cellular
membranes. The detailed results from simple model systems can be then
used to understand the role of ions in complex biological systems.

Zwitterionic phosphocholine (PC) lipid bilayers are commonly used model systems for
cellular membranes. Interactions of biological cations, especially Na$^+$ and
Ca$^{2+}$, with PC bilayers are widely studied with
experiments~\cite{akutsu81,altenbach84,seelig90,cevc90,tocanne90,binder02,pabst07,uhrikova08}
and classical MD simulations~\cite{bockmann03,bockmann04,Berkowitz12,melcrova16,javanainen17}.
The details of ion binding are, however, not agreed in the literature.
%not agreed in the literature \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Non-invansive spectroscopic methods, like nuclear magnetic resonance (NMR), scattering
and infrared scpectroscopy, give accurate information about ion binding
in PC lipid bilayers\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02,pabst07,uhrikova08}.
Interpretation of these experiments suggests that Na$^+$ ions have
negligible binding in PC lipid bilayers with submolar concentrations, while Ca$^{2+}$ specifically
binds to phosphate groups of two lipid molecules. Atomistic resolution
molecular dynamics (MD) simulation models, however, predict significantly
stronger binding for the cations and interactions with 3-4 lipids,
including also interactions with carbonyl
oxygens~\cite{bockmann03,bockmann04,melcrova16,javanainen17}.
Some experiments have also been interpreted to support the predictions from
MD simulations~\cite{bockmann03,vacha09a}.

Recent work published by the NMRlipids project (\url{nmrlipids.blogspot.fi})~\cite{catte16}
made an attempt to resolve the apparent controversies. The study presented 
a direct comparison between ion binding affinity between simulations and
experiments by using NMR data for lipid headgroup order 
parameters and the electrometer concept~\cite{seelig87}.
Using massive data collected by Open Collaboration method, it was concluded
that the accuracy of the current state of the art lipid models for MD
simulations is not sufficient for the detailed interpretation of the cation
interactions with PC lipid bilayers~\cite{catte16}.

%\todo{PAVEL: introduce previous theoretical work that discusses cation binding to POPC w.r.t. its specific moiteties, e.g. Lukas' paper}.

%While relative binding affinity of different ions in PC lipids is agreed to follow Hoffmeister
%series, the molecular details of binding and binding energetics are

@@

In this work we show that the cation binding behavior
in MD simulations of 1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer can be significantly 
improved by implicitly including electronic polarizability
in the polar region of lipid molecules. The electronic
polarizability is included by using the electronic
continuum correction (ECC)~\cite{leontyev11}, which
has been previously shown to improve the behaviour of
MD simulations of ions in bulk water~\cite{Jungwirth2017,Jungwirth2015,kohagen14,kohagen16}. 
As a starting point we use the parameters from Lipid14 model \cite{dickson14},
which gave the best cation binding behaviour in the previous study \cite{catte16}.
The developed ECC-lipid parameters reproduce the experimentally
measurable structural parameters of POPC lipid bilayer with the 
accuracy comparable to the other state of the art lipid models.

%We will validate such a model by comparing its lipid bilayer structure
%to x-ray scattering data and its head group structure to NMR experiments. 
%We will then quantify the accuracy of the head group order parameter response to the bound charge
%and compare it to the experimental NMR data of sodium and calcium binding in PC lipid bilayers. 
%In addition, we will quantify relative binding affinities of different moieties in POPC to such cations. 
%This will be discussed in terms of binding modes, stoichiometry and binding isotherms and their models used in experiments.

Since the cation binding affinity and the headgroup 
order parameter changes are in good agreement with experiments
in the proposed ECC-lipids, it can be used to interpret the
related structural changes and ion binding stoichiometry.
New lipid models with correct ion binding affinity in lipid
bilayers are necessary in applications of 
MD simulations with physiological salt conditions. 
The overestimated cation binding in the current lipid models \cite{catte16}
may lead to significant artefacts in MD simulations. For example,
articifially positively charged membranes overestimate interactions
with biomolecules having opposite sign.

%MD simualtions (e.g.~divalent cation-induced charge inversion of bacterial membranes~\cite{Luan2016}

%\todo{JOE: this may be a too daring statement. \\
%  SAMULI: I think that the divalent cation induced charge inversion is not an artefact,
%  but it might happen with too small concentrations. 
%  On the other hand, it is not a new observation either, see e.g. literature cited in NMRlipids IV project.}).

@@

\subsection{Electrometer concept} \label{section:electrometer}
Ion binding was compared between experiments and simulations by 
using lipid head group order parameters and the
''electrometer concept'' \cite{seelig87,catte16}.
The concept is based on the experimental observation that the C-H bond
order parameters of $\alpha$ and $\beta$ carbons in PC lipid head group
(see Fig.~\ref{simVSexpNOions}) are proportional to the amount 
of unit charge bound per lipid,  $X^\pm$~\cite{seelig87}.
Change in the order parameters measured with varying aqueous
ion concentration can be then related to the amount of bound ions.

The change of the head group order parameters is empirically quantified
as~\cite{seelig87,ferreira16}
\begin{equation}\label{OPchangeEQ}
  \Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) \approx \frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
$m_i$ is an empirical constant depending on the valency and position of the bound charge,
and the experimental value \cite{seelig77,davis83}, $\chi \approx$\,167\,kHz, is used for the quadrupole coupling constant.
Atomic absorption spectra and $^2$H NMR data gave
$m_\alpha=-20.5$~kHz  and $m_\beta=-10.0$~kHz for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
The slopes are negative, because recent analysis concluded 
that the order parameters decrease with bound positive charge 
and increase with bound negative charge when the signs are taken 
in account \cite{ollila16,catte16}. This is rationalized as a change
of lipid head group dipole tilt toward water phase with bound positive
charge and {\it vice versa} with negative charge \cite{seelig87}.

The concept can be used to compare the ion binding affinity in lipid 
bilayers between MD simulations and  NMR experiments, because the order
parameters can be accurately determined from both techniques~\cite{ollila16}.
The order parameters for all C-H bonds in lipid molecules, including
$\alpha$ and $\beta$ segments in head group, can be accurately measured
by using $^2$H NMR or $^{13}$C NMR techniques. From MD simulations the
order parameters can be calculated by using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and the average is taken over all sampled configurations \cite{ollila16}.

The measured order parameter change depends on the response of 
headgroup on bound charge, i.e. $m_i$ in Eq.~\ref{OPchangeEQ},
and the ion binding affinity. Thus, the former property has to be
well quantified before using using the electrometer concept to
analyze binding affinities. This is done experimentally for wide range
of systems~\cite{seelig87,beschiasvili91}. To calibrate 
the head group order parameter response also in simulations,
we use experimental data for dihexadecyldimethylammonium (C$_{12}$C$_{16}$$^+$N2C$_1$Br$^-$)
in POPC bilayer \cite{scherer89}. Dihexadecyldimethylammonium
is a cation surfactant having two acyl chains and bearing a unit
charge in the hydrophilic end. Thus, it is expected to locate
in bilayer similarly to the phospholipids and the molar ratio then 
gives directly the amount of bound unit charge per lipid $X^\pm$ in these systems \cite{scherer87}.

@@

In order to validate the newly developed model, ECC-lipids, 
we compared our simulation results without any ions to
NMR order parameters measurements and x-ray scattering form factors 
(Fig.~\ref{simVSexpNOions} and Table~\ref{tab:apls}). 
The tail order parameters being already highly optimized in the original Lipid14 model~\cite{dickson14}
are found to match the experimental values nicely.
\todo{The discussion about acyl chain and headgroup order parameters needs to be
  finished when the value for C2, the scale are fixed in the figure and
  size of the points are fixed in the figure.}
The headgroup order parameters are slightly larger in the ECC-lipid model,
which is in line with larger P-N vector angle in Fig.~\ref{OrderParameterCHANGESsurf}.
With the current data we cannot, however, conclude which is the more realistic
conformation since in both models other headgroup order parameter agrees
with experiments within error bars and other does not.
Slight changes are also observed in glycerol backbone order parameters,
bu the accuracy of both models is comparable to the state of art lipid models
available in literature \cite{ollila16}. The agreement between the x-ray scattering
form factors and the areas per molecule from simulations and experiments
confirm that the membrane structural properties are well captured. 
A structural comparison of ECC-lipids with Lipid14 can be found 
in SI along with results with other water models. 
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.

@@

The headgroup order parameter changes with increasing amount of 
cationic surfactant dihexadecyldimethylammonium bromide (C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) 
is compared between experiments~\cite{scherer89} and simulations in
Fig. \ref{OrderParameterCHANGESsurf}. 
The observed order parameter decrease is simulations and experiments
can be approximated to be linear at least with the mole fractions
below $\sim$30\%, as expected from Eq. \ref{OPchangeEQ}.
The slope is, however, too steep in Lipid14 model indicating that 
the head group order parameters are too sensitive to a bound charge.
The ECC-lipids model gives a slope in very good agreement with experiments
for the $\alpha$ segment, while the slope is slightly
underestimated for the $\beta$ segment.
%This has to be taken into account when interpreting the
%head group order parameter response to CaCl$_2$ concentration in Lipid14 model.
%observed in  Ref. \citenum{catte16}, see also Fig. \ref{OrderParameterCHANGESnewMODELS}.
%The present comparison of head group order parameter changes 
%suggests that the observed overestimated response of all models to 
%increasing CaCl$_2$ concentration in \cite{catte16} 
%can be explained at least in part by a high sensitivity of the head group order parameter response
%to the bound charge.
\todo{SAMULI: We could calculate the slopes from simulations, but I am not sure if
we would actually learn anything useful from this.}

The headgroup P-N vector angle with respect to the membrane normal
is also shown as a function of cationic surfactant mole fraction
in Fig.~\ref{OrderParameterCHANGESsurf}.
The headgroup orients more parallel to the membrane
normal with increasing amount of bound cations, as expected~\cite{seelig87}. 
The effect is more pronounced in Lipid14 than 
in ECC-lipids model, which is in line with the order parameter 
results and the reduced charge--dipole interactions in the ECC-lipid model.
The response of $\alpha$-order parameter to bound positive charge
in ECC-lipid model is in good agreement with experiments.
The model can be thus used to study changes of lipid P-N vector
in varying conditions.

%
%Therefore we use the data in Fig.~\ref{OrderParameterCHANGESsurf}
%to suggest a linear relation between $\alpha$-carbon order parameter
%and average P-N vector angle changes $\Delta \Theta_{\rm P-N}=?? \Delta S_{\rm CH}^{\alpha}$.
%This can be used to estimate the change in P-N vector angle
%from experiments measuring changes in $\alpha$-carbon order parameter. 
%\todo{SAMULI: I think that we should establish the relation between 
%order parameter and P-N vector angle changes by using ECC-lipids model.
%This would be useful for people measuring aplha order parameter
%changes from experiments.}
%
%For example, $\Delta S_{\rm CH}^{\alpha}=0.018$ measured due to the
%addition of 66.8 $\mu$M concentration of Mellittin in bulk water \cite{kuchinka89}
%gives $\Delta \Theta_{\rm P-N} = 2.5^{\circ}$.
%\todo{I looked this roughly from the figure, should be calculated from 
%the actual relation if we decide to put this.}

@@

The binding affinities are quantified by using
the ion density profiles along membrane normal shown
in Fig.~\ref{fig:cacl-dens}. The density profiles show
larger Ca$^{2+}$ density peak in lipid headgroup region for
Lipid14 model with Dang and ECC-ions than for the ECC-lipid model.
The relative surface excess calculated from Eq.~\ref{surfexcess} gives 
$\Gamma_i^w = 0.07 \pm 0.01 nm^{-2}$ for the ECC-lipid model,
which is significantly smaller than $\Gamma_i^w = 0.13 \pm 0.01 nm^{-2}$ for
Lipid14 with \AA{}qvist and $\Gamma_i^w = 0.3 \pm 0.03 nm^{-2}$ with Dang ions.

\todo{Below analysis is done in a stupid way to get some idea. I would fine it useful to
  do this analysis by using the density profiles, but it is not necessary.}
Rough estimates for the free energy difference between bound and unbound cation
are given by
\begin{equation}
  \Delta G={\rm k_b T}\log(\frac{p_o}{p_i}),
\end{equation}
where $p_o$ and $p_i$ are estimated from the Ca$^{2+}$densities in bulk water and
in the maximum density in bilayer, respectively. The density profiles in
Fig. \ref{fig:cacl-dens} give $\sim 0.8$ k$_{\rm b}$T for the free energy difference
between bound and unbound Ca$^{2+}$ ions in ECC-lipid model and $\sim 1.4$ k$_{\rm b}$T
in Lipid14 with \AA{}qvist.

@@

\begin{figure}[]
  \centering
  \includegraphics[height=9.0cm,angle=-90]{../Fig/bound-CAs_conc-eccl17.eps}
  \caption{\label{fig:cacl-bind}
    Binding isotherm assuming stoichiometry of 2~POPC:1~\ce{Ca^{2+}} as used in \cite{altenbach84} fits our simulation data well .
    }
\end{figure}

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=6.0cm]{../Fig/volmap_resid10_Ca_Cl_PO4Cent.png} \\
  \includegraphics[width=4.0cm]{../Fig/volmap_resid10_Ca_Cl_sn1Cent.png}
  \includegraphics[width=3.6cm]{../Fig/volmap_resid10_Ca_Cl_sn2Cent.png}
  \caption{\label{fig:volmaps}
      Contours of probability isodensities of \ce{Ca2+} with respect to 
      various moieties fixed in space (highlighted with transparent spheres): 
      phosphate moiety, side chain 1 carbonyl group and side chain 2 carbonyl group.
      Shown contours suggest that the dominant contribution 
      to \ce{Ca2+} binding likely comes from the phosphate group, 
      but remain inconclusive on the details of the interactions with any of the two carbonyl groups. 
  }
\end{figure}

@@
